# Logical vectors

## Simplifying predicates

[3.3.5. Simplifying predicates](https://deepr.gagolewski.com/chapter/130-logical.html#simplifying-predicates)

> Assuming that a, b, and c are numeric vectors, simplify the following expressions:

> !(b\>a & b\<c),

One of the sub-conditions has to be FALSE for the result to be TRUE:

!b\>a \| !b\<c

> !(a\>=b & b\>=c & a\>=c),

Same again:

!a\>=b \| !b\>=c \| !a\>=c

> a\>b & a\<c \| a\<c & a\>d,

I think I complicated rather than simplified, but:

a\<c & (a\<b \| a\>d)

> a\>b \| a\<=b,

Uhh, `TRUE`?

> a\<=b & a\>c \| a\>b & a\<=c,

Hmm, not sure this one can be simplified.

> a\<=b & (a\>c \| a\>b) & a\<=c,

a is less than or equal to b AND less than or equal to c, but must be greater than at least one of b OR c, so always: `FALSE`.

> !all(a \> b & b \< c).

If all(p) is equivalent to !any(!p) then isn't !all(p) equal to any(!p)?

any(!a\>b \| !b\<c)

## Nesting ifelse

[3.4. Choosing elements with ifelse](https://deepr.gagolewski.com/chapter/130-logical.html#choosing-elements-with-ifelse)

## `ifelse()` to generate Guassian mixtures

> we generated a variate from the normal distribution that has the expected value of −2 with probability 20%, and from the one with the expectation of 3 otherwise. Thus inspired, generate the Gaussian mixtures:

> $\frac{2}{3}X + \frac{1}{3}Y$, where $X ~ N(100, 16)$ and $Y ~ N(116, 8)$

```{r gauss-1}
n <- 10000

z <- ifelse(runif(n) <= (2/3), rnorm(n , 100, 16), rnorm(n, 116, 8))

hist(z, breaks = 101, probability = TRUE, main = "", col = "white")
```

> $0.3X + 0.4Y + 0.3Z$, where $X$ \~ $N(-10, 2)$, $Y$ \~ $N(0,2)$, and \$Z \$ \~ $N(10,2)$

```{r}
r <- runif(n)

d <- ifelse(
  r <= 0.2,
  rnorm(n, -10, 2),
  ifelse(r <= 0.7, rnorm(n, 0, 2), rnorm(n, 10, 2))
)

hist(d, breaks = 101, probability = TRUE, main = "", col = "white")
```

## Exercises

[3.5. Exercises](https://deepr.gagolewski.com/chapter/130-logical.html#exercises)

> *Answer the following questions.*
>
> *Why the statement "The Earth is flat or the smallpox vaccine is proven effective" is obviously true?*

Because at least one of those two statements is true, which is all you need with an OR.

> *What is the difference between `NA` and `NA_real_`?*

`NA` is of type logical wheras `NA_real_` is of type numeric, i.e. they indicate a missing value within those respective vector types.

> *Why is "FALSE & NA" equal to FALSE, but "TRUE & NA" is NA?*

As long as one side of an & is FALSE, the expression will evaluate to FALSE. With "TRUE & NA" the expression could evaluate to either TRUE or FALSE depending on what the NA value is.

> *Why has ifelse(x\>=0, sqrt(x), NA_real\_) a tendency to generate warnings and how to rewrite it so as to prevent that from happening?*

> *What is the interpretation of `mean(x >= 0 & x <= 1)`?*

This calculates the proportion of vector x which lies between 0 and 1. `mean` coerces the `logical` result of the expression to `numeric` i.e. `TRUE` to 1 and `FALSE` to 0.

> *For some integer x and y, how to verify whether 0\<x\<100, 0\<y\<100, and x\<y, all at the same time?*

`0<x & x<y & y<100`

> *Mathematically, for all real* $x,y>0$*, we have* $logxy=logx+logy$*. Why then `all(log(x*y) == log(x)+log(y))` can sometimes return `FALSE`? How to fix this?*

Differences can occur in the results due to over or under accuracy in calculations using the double-precision floating point system.

> *Is x/y/z always equal to x/(y/z)? How to fix this?*

```{r}
# 1 / 0.1
0.1 / 0.1 / 0.1
# 0.1 / 1
0.1 / (0.1 / 0.1)
```
Clearly not as a result of order of operations. I'm not sure what is meant by fixing this.

> *What is the purpose of very specific functions such as log1p and expm1 (see their help page) and many others listed in, e.g., the GNU GSL library \[28\]? Is our referring to them a violation of the beloved "do not multiply entities without necessity" rule?*

> *If we know that x may be subject to error, how to test whether x\>0 in a robust manner?*

We can for instance use `abs(x - 0) < e-8` to check for within an assumed error margin or more simply round `round(x) > 0)`.

> *Is "y\<-5" the same as "y \<- 5" or rather "y \< -5"?*

It's the same as the former, evaluated as assignment rather than less than negative 5.

## Equality functions
> *What is the difference between `all` and `isTRUE`? What about `==`, `identical`, and `all.equal`? Is the last one properly vectorised?*

`all` checks if every element of a vector is TRUE, including coercion to TRUE, and `isTRUE` checks if the value passed to it is equal to the value `TRUE`.

```{r isTrue}
# Are all values TRUE?
all(c(TRUE, TRUE))

# Is it a TRUE value?
isTRUE(c(TRUE, TRUE))
```

`==` checks for equality between each respective element of two vectors, return a result for each pair. `identical` checks if two objects are exactly equal, a single result for whether the entire thing is the same. `all.equal()` is similar to identical, but instead tests for near equality.

```{r all.equal}
c(1, 2, 3) == c(1, 2, 3)
identical(c(1, 2, 3), c(1, 2, 3))

identical(c(1, 2, 3), c(1.1, 2.1, 3.1))
all.equal(c(1, 2, 3), c(1.1, 2.1, 3.1), tolerance = 0.1)

```
