# Logical vectors

## Simplifying predicates

[3.3.5. Simplifying predicates](https://deepr.gagolewski.com/chapter/130-logical.html#simplifying-predicates)

> Assuming that a, b, and c are numeric vectors, simplify the following expressions:

> !(b\>a & b\<c),

One of the sub-conditions has to be FALSE for the result to be TRUE:

!b\>a \| !b\<c

> !(a\>=b & b\>=c & a\>=c),

Same again:

!a\>=b \| !b\>=c \| !a\>=c

> a\>b & a\<c \| a\<c & a\>d,

I think I complicated rather than simplified, but:

a\<c & (a\<b \| a\>d)

> a\>b \| a\<=b,

Uhh, `TRUE`?

> a\<=b & a\>c \| a\>b & a\<=c,

Hmm, not sure this one can be simplified.

> a\<=b & (a\>c \| a\>b) & a\<=c,

a is less than or equal to b AND less than or equal to c, but must be greater than at least one of b OR c, so always: `FALSE`.

> !all(a \> b & b \< c).

If all(p) is equivalent to !any(!p) then isn't !all(p) equal to any(!p)?

any(!a\>b \| !b\<c)

## Nesting ifelse

[3.4. Choosing elements with ifelse](https://deepr.gagolewski.com/chapter/130-logical.html#choosing-elements-with-ifelse)

## `ifelse()` to generate Guassian mixtures

> we generated a variate from the normal distribution that has the expected value of −2 with probability 20%, and from the one with the expectation of 3 otherwise. Thus inspired, generate the Gaussian mixtures:

> $\frac{2}{3}X + \frac{1}{3}Y$, where $X ~ N(100, 16)$ and $Y ~ N(116, 8)$

```{r gauss-1}
n <- 10000

z <- ifelse(runif(n) <= (2/3), rnorm(n , 100, 16), rnorm(n, 116, 8))

hist(z, breaks = 101, probability = TRUE, main = "", col = "white")
```

> $0.3X + 0.4Y + 0.3Z$, where $X$ ~ $N(-10, 2)$, $Y$ ~ $N(0,2)$, and $Z $ ~ $N(10,2)$

```{r}
r <- runif(n)

d <- ifelse(
  r <= 0.2,
  rnorm(n, -10, 2),
  ifelse(r <= 0.7, rnorm(n, 0, 2), rnorm(n, 10, 2))
)

hist(d, breaks = 101, probability = TRUE, main = "", col = "white")
```

